Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-06-20 paper.tex
XCR update

\C{A quantity that is of great scientific interest is the CR pressure
  ($P_\CR$) relative to the thermal pressure ($P_\rmn{th}$) as it
  directly bias galaxy cluster observables and the resulting
  cosmological constraints, such as the hydrostatic mass bias or the
  Sunyaev-Zel’dovich signal measured by Planck, South Pole Telescope,
  or Atacama Cosmology Telescope. In the following we estimate the
  mass scaling of the relative CR pressure, $X_\CR =
  P_\CR/P_\rmn{th}$, using the CR formalism outlined in
  \cite{2007A&A...473...41E,2010MNRAS.409..449P}. We derive the CR
  pressure from the CR distribution function $f = C\,p^{-\alpha}$,
  where $C$ is the normalization of each power-law, $p=P/m_\rmn{p} c$
  is the normalized CR proton momentum, and $\alpha$ is the CR
  spectral index. The pressure is given by $P_\CR \propto
  C=\tilde{C} \,\rho_\rmn{gas}$, where $\tilde{C}\propto M^{0.44}$ for
  clusters with a mass $\mvir\gtrsim 10^{14}\,\msun$
  \cite{2010MNRAS.409..449P}. We determine the mass scaling of the gas
  density through its relation to the enclosed gas fraction
  $f_\rmn{gas}$, and find that $\rho_\rmn{gas}\propto f_\rmn{gas}
  \propto M^{0.135}$ \cite{2009ApJ...693.1142S}. The thermal pressure
  is derived from the temperature in virial equilibrium, and scales as
  $P_\rmn{th}\propto k_\B T \propto G M/R \propto M^{0.67}$. Finally
  we can derive the mass scaling of the volume weighted relative CR
  pressure from
\begin{equation}
 <X_\CR> = \frac{<P_\CR>}{<P_\rmn{th}>} \propto \frac{<\rho_\rmn{gas}\,
   \tilde{C}>}{<k_\B T>} \propto M^{-0.1}\,,
\end{equation}
which is roughly a constant, i.e. almost independent on mass. Hence we
adopt a constant $<X_\CR>\approx 0.02$ derived from a representative
simulated cluster \cite{2008MNRAS.385.1211P}. Since the CRs that build
up the bulk of the CR pressure also give rise to gamma-ray emission
through the production of decaying pions, we use gamma-ray upper
limits from Fermi-LAT to constrain the volume weighted $X_\CR$. In
Fig.~\ref{fig:XCR} we show the volume weighted relative CR pressure
for 32 clusters, for which LAT upper limits that assume King profiles
are derived. For most of the clusters in the sample we can constrain
$X_\CR < (0.1-0.2)$ using the extended gamma-ray upper limits, where
the most constraining clusters are Norma and Coma with $X_\CR <
0.07$. Most of the 32 clusters have $X_\CR$ upper limits within a
factor few from the best constraining clusters, which suggests that
stacking of clusters will be able to improve these limits. This might
even be improved further using a scaled stacking with our model as a
prior. In the recent work \cite{2010ApJ...717L..71A} where the 1.5
year Fermi data was used to derive limits on $X_\CR$, they found that
the most massive clusters are on average more constraining, while we
find better limits on the low mass clusters. This difference is
explained by the different priors of $P_\CR$; they assumed a mass
independent CR pressure derived from a single CR power-law
distribution function, while we adopt a prior infered from cosmological
simulations that dynamically trace the CRs during cluster assembly and
allow for a varyning spectral index where $P_\CR\sim
M^{0.44}$. Furthermore, the gamma-ray flux predictions that we use in
this work to constrain $X_\CR$ assumes the true gas density
distribution for each cluster \footnote{Derived from the parameters in
  the extended HIFLUGCS catalogue.} that accounts for
cluster-to-cluster scatter, in contrast to the mass-to-luminosity
scaling relations that was used in previous work.}

>>>>

\C{A quantity that is of great scientific interest is the CR pressure
  ($P_\CR$) relative to the thermal pressure ($P_\rmn{th}$) as it
  directly bias galaxy cluster observables and the resulting
  cosmological constraints, such as the hydrostatic mass bias or the
  Sunyaev-Zel’dovich signal measured by Planck, South Pole Telescope,
  or Atacama Cosmology Telescope. In the following we estimate the
  mass scaling of the relative CR pressure, $X_\CR =
  P_\CR/P_\rmn{th}$, using the CR formalism outlined in
  \cite{2007A&A...473...41E,2010MNRAS.409..449P}. We derive the CR
  pressure from the CR distribution function $f = C\,p^{-\alpha}$,
  where $C$ is the normalization of each power-law, $p=P/m_\rmn{p} c$
  is the normalized CR proton momentum, and $\alpha$ is the CR
  spectral index. The pressure is given by $P_\CR \propto
  C=\tilde{C} \,\rho_\rmn{gas}$, where $\tilde{C}\propto M^{0.44}$ for
  clusters with a mass $\mvir\gtrsim 10^{14}\,\msun$
  \cite{2010MNRAS.409..449P}. We determine the mass scaling of the gas
  density through its relation to the enclosed gas fraction
  $f_\rmn{gas}$, and find that $\rho_\rmn{gas}\propto f_\rmn{gas}
  \propto M^{0.135}$ \cite{2009ApJ...693.1142S}. The thermal pressure
  is derived from the temperature in virial equilibrium, and scales as
  $P_\rmn{th}\propto k_\B T \propto G M/R \propto M^{0.67}$. Finally
  we can derive the mass scaling of the volume weighted relative CR
  pressure from
\begin{equation}
 <X_\CR> = \frac{<P_\CR>}{<P_\rmn{th}>} \propto \frac{<\rho_\rmn{gas}\,
   \tilde{C}>}{<k_\B T>} \propto M^{-0.1}\,,
\end{equation}
which is roughly a constant, i.e. almost independent on halo
mass. This suggests that CRs are acclerated at the strongest shocks,
in the early universe or between voids and pre-collapsed IGM, and then
experience a similar transport history in the form of adiabatic
compression or similarily weaker shocks. Since the CRs that build up
the bulk of the CR pressure also give rise to gamma-ray emission
through the production of decaying pions, we use gamma-ray upper
limits from Fermi-LAT to constrain $X_\CR$. In Fig.~\ref{fig:XCR} we
show the volume weighted relative CR pressure for 32 clusters, for
which LAT upper limits that assume King profiles are derived. We adopt
a constant $<X_\CR>\approx 0.02$ derived from a representative
simulated cluster \cite{2008MNRAS.385.1211P} and find that for most of
the clusters in the sample we can constrain $X_\CR < (0.1-0.2)$ using
the extended gamma-ray upper limits, where the most constraining
clusters are Norma and Coma with $X_\CR < 0.07$. Most of the 32
clusters have $X_\CR$ upper limits within a factor few from the best
constraining clusters, which suggests that stacking of clusters will
be able to improve these limits. This might even be improved further
using a scaled stacking with our model as a prior. In a recent work
\cite{2010ApJ...717L..71A}, where the 1.5 year Fermi data was used to
derive limits on $X_\CR$, they found that the most massive clusters
are on average more constraining, while we find better limits on the
low mass clusters. This difference is explained by the different
priors on $P_\CR$; they assumed a mass independent CR pressure derived
from a single CR power-law distribution function, while we adopt a
prior inferred from cosmological simulations that dynamically trace the
CRs during cluster assembly and allow for a varying spectral index
where $P_\CR\sim M^{0.44}$. Furthermore, the gamma-ray flux
predictions that we use in this work to constrain $X_\CR$ assumes the
true gas density distribution for each cluster \footnote{Derived from
  the parameters in the extended HIFLUGCS catalogue.} that accounts
for cluster-to-cluster scatter, in contrast to the mass-to-luminosity
scaling relations that was used in previous work.}
